# Chapter 3: Random vectors

## Random variable

## Cumulative distribution function

## Joint distribution
*A particle can have a position or a velocity, but strictly speaking it cannot hace both... The more we clarify the secret of position, the more deeply the secret of velocity hides... We cna distribute uncertainty as we wish, but we can never eliminate it.*

### Joint Distribution Function
Throughout this unit we assume a probability sapce $(\Omega,\mathcal{F},P)$ correspondig to a given random experiment.

Let us establish some **Notation**: Given random variables
$$
\begin{align}
X \colon \Omega \longrightarrow
 \mathbb{R}
\quad
X_1,\cdots, X_n \colon \Omega \longrightarrow
 \mathbb{R}
 \end{align}
$$

and subsets $B,B_1,\dots,B_n\subset\mathbb R$, we write:

$$
[ X \in B] \space \colon= \{\omega \in \Omega \colon X(\omega) \in B, \}
$$

$$
[X_1 \in B_1, \cdots, X_n \in B_n] \space \colon = \bigcap_{k=1}^{n}\{\omega \in \Omega \colon X_k(\omega) \in B_k\}
$$
and, for $A\subset\mathbb R^{n}$,

$$
[(X_1,\cdots,X_n) \in A] \space \colon = \{\omega \in \Omega \colon (X_1(\omega),\cdots,X_n(\omega))\in A\}
$$


Recall from the previous unit that all probabilistic information about a single random variable $X$ is contained in its distribution function, since, having the latter, one can obtain the probability of any event whose occurrence or non-ocurrence depends on the value taken by $X$.

Two random variables may be different when seen as functions defined over the sample space $\Omega$, but be identical on terms of their distribution, and thus, from the probabilistic point of view, they give us exactly the same information and can therefore be used interchangeably for the same purpose.

**Example - Symmetry of the uniform distribution $(0,1)$**
Suppose we conduct a random experiment where we select a real number at random in the interval $(0,1)$. Let us denote this quantity as a random variable $X$. Mathematically, we say:

$$
X \sim Uniform(0,1).
$$
Now define a new random variable:
$$
Y=1-X
$$

**What happens with the distribution of $Y$**
Although $X$ and $Y$ are different functions defined on the same sample space -in fact, $X(x) =x$ and $Y(x)=1-x$ - they share a very special property: They have the **same distribution**.

This property is due to the simmetry of the continuous uniform distribution on $(0,1)$ with respect to the point $x=0.5$.

We can verify this idea via simulation:

::: {.callout-note title="$Uniform(0,1)$ simulation of $X$ and $Y$"}
#### $Uniform(0,1) in R & Python
::: {.panel-tabset}
### R
```{r}
#| label: Symm-unif-r
#| code-fold: true

set.seed(123)           # Reproducibility
n <- 10000              # Number of simulations
x <- runif(n)           # X ~ Uniform(0,1)
y <- 1 - x              # Y = 1 - X

hist(x, breaks = 30, freq = FALSE,
     col = rgb(1,0,0,0.5), main = "Distribution of X and Y = 1 - X", xlab = "")
hist(y, breaks = 30, freq = FALSE,
     col = rgb(0,0,1,0.5), add = TRUE)
legend("topright", legend = c("X", "Y = 1 - X"),
       fill = c(rgb(1,0,0,0.5), rgb(0,0,1,0.5)))
```
### Python
```{python}
#| label: sym-unif-py
#| code-fold: true

import numpy as np
import matplotlib.pyplot as plt

# Reproducibility
np.random.seed(123)

# Number of simulations
n = 10000

# X ~ Uniform(0,1)
x = np.random.uniform(0, 1, n)

# Y = 1 - X
y = 1 - x

# Histogram of X
plt.hist(x,
         bins=30,          # Number of bins
         density=True,     # relative frequency
         alpha=0.5,
         color='red',
         label='X')

# Histogram of Y
plt.hist(y,
         bins=30,
         density=True,
         alpha=0.5,
         color='blue',
         label='Y = 1 - X')

# 
plt.title('Distribution of X and Y = 1 - X')
plt.xlabel('Value')
plt.ylabel('Density')
plt.legend()
plt.tight_layout()
plt.show()
```

:::
:::

The histogram almos perfectly overlap, illustrating that $X$ and $Y$ are different as functions, they have the same distribution. That is:

$$
X \stackrel{d}{=} Y
$$

This example helps us understand the difference between:

- **Random variables as functions**
- **Probability distributions**

it also shows how a simple transformation can generate a new variable with identical probabilistic properties.

**Example: Simulating Exponential Variables from Uniforms**

Suppose we want to generate $n$ random numbers that can be regarded as realization of a random variable $Y$ with exponential distribution parameter $\lambda=1$:

$$
Y \sim Exp(1)
$$

*Method 1 - Classic inverse transform.* Generates $x_k\sim\mathrm{Uniform}(0,1)$, then:

$$
y_k = -ln(x_k), \quad k=1,\cdots, n.
$$
follows an exponential distribution with parameter $\lambda=1$. This transformation is based on the inverse of the exponential cumulative distribution function.

*Method 2 - Alternative Equivalent Trasnformation*
We can also use:

$$
z_k = -ln(1-x_k), \quad k=1, \cdots, n
$$
Since $1-x_k \sim Uniform(0,1)$, this transformation also yields exponential variables.

Next, we ran a simulation to compare the two transformations.

::: {.callout-note title="Exponential transformation"}
#### Exponential transformation in R & Python
::: {.panel-tabset}
### R
```{r}
#| label: exp-transform-r
#| code-fold: true

set.seed(42)
 n <- 10000
 x <- runif(n)
 y <- -log(x)
 z <- -log(1 - x)

hist(y, breaks = 50, freq = FALSE,
     col = rgb(1,0,0,0.4), main = "y = -log(x) vs z = -log(1 - x)")
hist(z, breaks = 50, freq = FALSE,
     col = rgb(0,0,1,0.4), add = TRUE)
curve(dexp(x, rate = 1), col = "black", lwd = 2, add = TRUE)
legend("topright", legend = c("y = -log(x)", "z = -log(1 - x)", "Exp(1) density"),
       fill = c(rgb(1,0,0,0.4), rgb(0,0,1,0.4), NA), border = NA,
       lty = c(NA, NA, 1), col = c(NA, NA, "black"))
```
### Python
```{python}
#| label: exp-transform-py
#| code-fold: true
import numpy as np
import matplotlib.pyplot as plt

# Reproducibility
np.random.seed(42)

# Number of simulations
n = 10_000

# X ~ Uniform(0,1)
x = np.random.uniform(0, 1, n)

# Transforms
y = -np.log(x)        # y = -log(x)
z = -np.log(1 - x)    # z = -log(1 - x)

# Histogram of y
plt.hist(y,
         bins=50,
         density=True,
         alpha=0.4,
         color='red',
         label='y = -log(x)')

# Histogram of z
plt.hist(z,
         bins=50,
         density=True,
         alpha=0.4,
         color='blue',
         label='z = -log(1 - x)')

# Theoric curve Exp(1): f(t) = e^{-t},  t ≥ 0
t = np.linspace(0, max(y.max(), z.max()), 400)
plt.plot(t, np.exp(-t), 'k-', linewidth=2, label='Exp(1) density')

# Graph
plt.title('y = -log(x) vs z = -log(1 - x)')
plt.xlabel('Value')
plt.ylabel('Density')
plt.legend()
plt.tight_layout()
plt.show()

```

:::
:::

Both transformations produce samples that closely follow the theoretical exponential density. This shows that multiple generating functions can lead to the same distribution.

When a problem involves several random variables, knowing their marginal distribution is **not enough** to fully describe the phenomenon. it is also necessary to undestand how they relate to each other -Their **dependence struscture**, which can be captured via:

- The join distribution;
- measure such as **covariance** or **correlation**;
- or more advanced tools such as **copulas**.

Let us illustrate this point with two examples.

**Examples of random variables: Independence and dependence**

**Example 1.1: Random Point inside the Square**

Consider the random experiment of selecting a point ar random inside the square with vertices $(0,0)$,$(1,0)$,$(1,1)$,$(0,1)$. Define the random variable $X$ and $Y$ as the x- and y-coordinates of the selected point, respectively.

in this type of experiment, the probability that the selected point lies in a subset $A \subset \mathbb{R}^{2}$ equals the ratio of the area of $A\cap C$ to the area of $C$, where $C$ is the square.

**Marginal Distributions:**
$$
F_{X}(x) = \begin{cases}
0 & if & x\leq 0 \\
x & if & 0<x<1 \\
1 & if & x\geq 1
\end{cases}
\qquad
F_{Y}(y)=\begin{cases}
0 & if & y \leq 0\\
y & if & 0 <y<1\\
1 & if & y\geq 1
\end{cases}
$$

**Observation:** Knowing the value of $X$ tells us nothing about $Y$, since $Y$ can range across the entire interval $(0,1)$. That is, $X$ and $Y$ are **indeoendent.**

**Example 1.2: Random Point on the Diagonal**
Now consider the experiment of selecting a point at random on the diagonal of slope 1 within the same square. Again, define $X$ and $Y$ as the x- and y-coordinates of the selected point.

In the case, the probability that the point lies in a subset  $A\subset\mathbb{R}^{2}$ equals the restio of the length of $A\cap D$ to the lenght of $D$, where $D$ is the diagonal.

**Marginal Distributions:**
$$
F_{X}(x) = \begin{cases}
0 & if & x\leq 0 \\
x & if & 0<x<1 \\
1 & if & x\geq 1
\end{cases}
\qquad
F_{Y}(y)=\begin{cases}
0 & if & y \leq 0\\
y & if & 0 <y<1\\
1 & if & y\geq 1
\end{cases}
$$

**Observation:** Although the marginal distribution function are exactly the same as in Example 1.1, the relationship between the variables is different: Here, knowing the value of $X$ immediately determines the value of $Y$, since they are equal: $Y=X$. Therefore, $X$ and $Y$ are **dependent.**

**Conceptual comparison**
- In the **Example 1.1**, the selection is made over the entire square. The variables $X$ and $Y$ exhibit **independence** behavior: the joint probability is the product of the marginals
- In **Example 1.2**, the selection is restricted to a line. Knowing one coordinate completely determines the other: there is **functional dependence**

| Characteristics | Example 1.1(Square) | Example 1.2 (Diagonal) |
|---           |:---:         |---:          |
| Selected region      | Square Area      | Diagonal line      |
| Relation between $X$ and $Y$      | independent      | dependent      |
| Marginal distribution      | $Uniform(0,1)$      |$Uniform(0,1)$    |
| Joint distribution      | Product of marginals      | Support over the line $y=x$      |

**Joint Distribution of a family of random variables**
For a single random variable, the distribution function $F_{X}(x)=P(X \leq x)$ plays a central role, as it encapsulates all the probabilistic information about that variable.

However, when dealing with a family of $n$ random variables $X_{1},X_{2}, \cdot, X_{n}$,. it is not enough to consider separately the marginal distribution functions of each:

$$
F_{X_{i}}(x_{i}) = P(X_{i} \leq x_i), \quad i=1,2,\cdots,n
$$

These functions describe the individual behavior of each variable but contain no information about how the variables relate to each other.

**Example:** It may happen that two random variables, the central object of study is the **joint distribution function**, defined as follows:

**Definition (Joint distribution function) (por numerar)**
Let $X_1,X_2,\cdots,X_n$ be n random variables. The function $F_{X_1,X_2,\cdots,X_n}\colon \mathbb{R}^{n} \rightarrow [0,1]$, defined by:

$$
F_{$X_1,X_2,\cdots,X_n$}($x_1,x_2,\cdots,x_n$)=P[X_1\leq x_1, \cdots X_n \leq x_n]
$$
is called the **joint distribution function** of $$X_1,X_2,\cdots,X_n$.

Let us illustrate this definition by finding the joint distribution function of $X$ and $Y$ in each of the two examples previously mentioned.

**Proposition (por numerar)**
Let $X$ and $Y$ be any two random variables, and let $F_{X,Y}$ be their joint distribution function. Then:
1. $\lim_{x\rightarrow \infty} F_{X,Y}(x,y) = F_{Y}(y) \quad \forall y \in \mathbb{R}$.

2. $\lim_{y\rightarrow \infty} F_{X,Y}(x,y) = F_{X}(x) \quad \forall x \in \mathbb{R}$.
**proof**
Let $y\in \mathbb{R}$, and let $(x_n)$ be a monotonically increasing sequence of real numbers such that $\lim_{n\rightarrow \infty} x_n = \infty$. then, the sequence of events $[X\leq x_n, Y \leq y]$ is non-decreasing, and:

$$
[Y \leq y]=\bigcup_{n=1}^{\infty}[X\leq x_n, Y\leq y]
$$
So:
$$
F_{Y}(y)=P[Y\leq y] = \lim_{n\rightarrow \infty}P[X\leq x_n, Y \leq y] = \lim_{n \rightarrow \infty}F_{X,Y}(x_n,y)
$$
the second identity is proved similarly.

This result extends to the case of $n$ random variables, just like other properties of the distribution function of a single variable. These properties, whose proofs are left as exercises, are started below:

**Proposition (por numerar)**
Let $X_1,X_2,\cdots X_n$ be $n$ random variables, and let $F_{X_1,X_2,\cdots X_n}$ be their joint distribution function. Then, for each $(x_1,x_2, \cdots x_n) \in \mathbb{R}^{n-1}$, the following hold:
1. The function
$$
x \mapsto F_{X_1,X_2,\cdots X_n}(x_1,\cdots, x_{j-1},x,x{j+1},\cdots x_n)
$$
is non-decreasing and right-continuous.

2. $\lim_{x\rightarrow \infty}F_{X_1,X_2,\cdots X_n}(x_1,\cdots, x_{j-1},x,x{j+1},\cdots x_n) = F_{X_1,\cdots, X_{j-1},X{j+1},\cdots X_n}(x_1,\cdots, x_{j-1},x,x{j+1},\cdots x_n)$

3. $\lim_{x \rightarrow -\infty} F_{X_1,\cdots X_n}(x_1,\cdots, x_{j-1},x,x{j+1},\cdots x_n)=0$

These conditions are necessary, but **not sufficient**, for a function $F$ to be a joint distribution function.

In reality, a distribution function represents a **measure**. in the case of a single variable, it corresponds to a measure over subsets of real numbers. In the case of the distribution function of two random variables $X$ and $Y$, it corresponds to a measure over subsets of $\mathbb{R}^{2}$.

in such case, this measure begins by assigning the value
$$
F_{X,Y}(x,y) = P[X\leq x, Y\leq y]
$$
to the infinite rectangle $(-\infty, x] \times (-\infty,y]$. More generally, if $x_1 \leq x_2$ and $y_1\leq y_2$, and $C$ is the rectangle $(x_1, x_2] \times (y_1,y_2]$, then:

$$
P[(X,Y) \in C] = F_{X,Y}(x_2,y_2) - F_{X,Y}(x_1,y_2) - F_{X,Y}(x_2,y_1) + F_{X,Y}(x_1,y_1)
$$
Thus, this value would be the measure assigned to rectangle $C$.

Note in particular, the quantity
$$
F_{X,Y}(x_2,y_2) - F_{X,Y}(x_1,y_2) - F_{X,Y}(x_2,y_1) + F_{X,Y}(x_1,y_1)
$$
is non-negative regardless of the joint distribution function $F_{X,Y}$.

it can be shown that this additional condition is sufficient for a function of two variables to represent the **joint distribution** of two random variables. That is, we have the following result:

**Proposition (por numerar)**
A function $F\colon \mathbb{R}^{2} \rightarrow \mathbb{R}$ represents the joint distribution function of a pair of random variables $X,Y$ if and only if the following conditions are satisfied:
1. For each $y\in \mathbb{R}$, the function $x \mapsto F(x,y) is non-decreasing, right-continuous, and
$$
\lim_{x\rightarrow -\infty}F(x,y)=0
$$

2. For each $x\in \mathbb{R}$, the function $y \mapsto F(x,y)$ is non decreasing, right-continuous, and
$$
\lim_{y\rightarrow -\infty}F(X,Y)= 0
$$

3. The functions $G\colon \mathbb{R} \rightarrow [0,1]$ and $H\colon \mathbb{R} \rightarrow [0,1]$, defined by:
$$
G(y) = \lim_{x\rightarrow \infty}F(x,y), \quad H(x)=\lim_{y\rightarrow \infty}F(x,y)
$$
are valid distribution functions of a single variable.

4. if $x_1\leq x_2$ and $y_1 \leq y_2$, then:
$$
F(x_2,y_2) - F(x_1,y_2) - F(x_2,y_1) + F(x_1,y_1) \geq 0
$$ 

Note that in the example considered above, we have:
$$
F(1, 1) − F(0, 1) − F(1, 0) + F(0, 0) = −1
$$
This means that if $F$ were the joint distribution function of a pair of random variables $X,Y$ and $C$ were the square $0< x \leq 1, 0<y\leq 1$ then:
$$
P[(X,Y) \in C] = -1
$$
which is a contradiction.

In general, the joint distribution function of $n$ random variables represents a **measure on** $\mathbb{R}^{n}$. The family of random variables $X_1,X_2, \cdots X_n$ can be viewed as a function from $\Omega$ into $\mathbb{R}^{n}$ that assigns to each $\omega \in \Omega$ the vector $(X_1(\omega),X_2(\omega), \cdots, X_n(\omega))$; in this way, we can say that the random variables form a **random vector** $(X_1,X_2, \cdots, X_n)$.

### Joint density functions
When studying the distribution of a single random variable, there are two cases in which the variable is determined by a density function: the discrete case and the absolutely continuous case.(Habría que definir
que es absolutamente continua en el caso uni variado, si no poner la definición
como nota al pie acá.)
This situation can be extended to a family of random variables, which is developed below. For clarity of exposition, we first address the case of a family consisting of two random variables, and then state the results for the general case.

**Definition (Bivariate discrete random vector)(por numerar)**
A pair of random variables $X,Y$ is said to form a **discrete random vector** if there exists a finite or countably infinite collection of vectors $(x_1, y_1), (x_2,y_2),\cdots$ such that:

1. $P[X=x_m, Y=y_m] > 0 \quad \forall m$
2. $\sum_{m}P[X=x_m, Y=y_m] = 1$

in this case, the function $f_{X,Y}\colon \mathbb{R}^{2} \rightarrow [0,1]$, defined by
$$
f_{X,Y}(x,y) = P[X=x,Y=y]
$$
is called the **joint density function** of the vector $(X,Y)$.
The property of **countable additivity** implies the following relationship:
$$
F_{X,Y}(x,y) = \sum_{(u,v)\in \mathbb{R}^{2} u \leq x, v\leq y} f_{X,Y}(u,v)
$$
More generally, if $A \subset \mathbb{R}^{2}$, the same property yields:

$$
\begin{equation}
P[(X,Y) \in A] = \sum_{\{(x,y)\in\mathbb{R}^2\,:\,(x,y)\in A\}} f_{X,Y}(x,y)
\end{equation}
$$

This expression is very useful for computing probabilities of events whose occurence depends on the value of both $X$ and $Y$.

**Example 1.9 (por numerar)**
Let $(X,Y)$ be a discrete random vector with joint density function given by:

$$
f_{X,Y}(x,y) = \begin{cases}
cx & if \quad x,y \in \{1,2,\cdots n\}\\
0 & otherwise
\end{cases}
$$

where $N$ is a positive integer. Find:
1. the value of $c$
2. $P[X=Y]$
3. $P[X<Y]$
4. $P[X>Y]$

**Solution**
1. 
$$
\sum_{x=1}^{N} \sum_{y=1}^{N} f_{X,Y}(x, y) = \sum_{x=1}^{N} \sum_{y=1}^{N} cx = c \sum_{x=1}^{N} \sum_{y=1}^{N} x = cN \sum_{x=1}^{N} x = cN \cdot \frac{N(N+1)}{2}
$$
Thus
$$
c = \frac{2}{N^{2}(N+1)}
$$


2. 
$$
P[X = Y] = \sum_{x=1}^{N} f_{X,Y}(x, x) = \sum_{x=1}^{N} cx = c \sum_{x=1}^{N} x = \frac{2}{N^2(N+1)} \cdot \frac{N(N+1)}{2} = \frac{1}{N}
$$

3. 
$$
P[X < Y] = \sum_{x=1}^{N-1} \sum_{y=x+1}^{N} f_{X,Y}(x, y) = \sum_{x=1}^{N-1} \sum_{y=x+1}^{N} cx = c \sum_{x=1}^{N-1} x(N - x)
$$
We simplify the last sum:
$$
= c \left[ N \sum_{x=1}^{N-1} x - \sum_{x=1}^{N-1} x^2 \right]
$$
Using lnown formulas:
$$
\sum_{x=1}^{N-1} x = \frac{(N-1)N}{2} \quad \sum_{x=1}^{N-1} x^2 = \frac{N(N-1)(2N-1)}{6}
$$
Thus.
$$
P[X<Y] = \frac{N-1}{3N}   
$$

4. 
$$
P[X > Y] = 1 - P[X = Y] - P[X < Y] = 1 - \frac{1}{N} - \frac{N - 1}{3N} = \frac{2(N - 1)}{3N}
$$

Note that in the type of problems illustrated by the previous example, one generally needs to compute the value of a double summation. This value can be obtained either by fixing $x$; or by fixing $y$ first and summing over the corresponding values of $x$, then completing the summation over $y$.

This is equivalent to using one of the following two methods:
$$
\begin{equation}
P[(X,Y) \in A] = \sum_x P[(X,Y) \in A, X=x] = \sum_x P[(x,Y)\in A, X=x] = \sum_x \sum_{y \colon (x,y) \in A} P[X=x,Y=y]
\end{equation}
$$

$$
\begin{equation}
P[(X,Y) \in A] = \sum_y P[(X,Y) \in A,Y=y]=\sum_y P[(X,y) \in A, Y=y] = \sum_y \sum_{x \colon (x,y) \in A}P[X=x,Y=y]
\end{equation}
$$

::: {.callout-note title="Example 1.9"}
#### Example 1.9 
::: {.panel-tabset}
### R
```{r}
#| label: cx-example-r
#| code-fold: true

# Parameters
N <- 10
c <- 2 / (N^2 * (N + 1))

# Create grid of points
grid <- expand.grid(x = 1:N, y = 1:N)
grid$f <- c * grid$x

# Classify region
grid$region <- with(grid, ifelse(x < y, "X < Y",
                          ifelse(x == y, "X = Y", "X > Y")))

# Color palette by region
region_colors <- c("X < Y" = "blue",
                   "X = Y" = "green",
                   "X > Y" = "red")

# Load ggplot2
library(ggplot2)

# Plot
ggplot(grid, aes(x = x, y = y)) +
  geom_tile(aes(fill = f), alpha = 0.4) +
  geom_point(aes(color = region), size = 4) +
  scale_color_manual(values = region_colors) +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(x = "x", y = "y", fill = "f(x,y)", color = "Region") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```
### Python
```{python}
#| label: cx-example-py
#| code-fold: true
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
from matplotlib.lines import Line2D

# Parámetros
N = 10
c = 2 / (N**2 * (N + 1))

# Crear grilla de puntos
x_vals = np.arange(1, N + 1)
y_vals = np.arange(1, N + 1)
grid = pd.DataFrame([(x, y) for x in x_vals for y in y_vals], columns=["x", "y"])
grid["f"] = c * grid["x"]

# Clasificar regiones
grid["region"] = np.where(grid["x"] < grid["y"], "X < Y",
                   np.where(grid["x"] == grid["y"], "X = Y", "X > Y"))

# Asignar colores
region_colors = {"X < Y": "blue", "X = Y": "green", "X > Y": "red"}
grid["region_color"] = grid["region"].map(region_colors)

# Crear figura
fig, ax = plt.subplots(figsize=(8, 8))

# Dibujar fondo con intensidad f(x,y)
pivot_f = grid.pivot(index="y", columns="x", values="f")
sns.heatmap(pivot_f, cmap="Reds", cbar_kws={'label': 'f(x,y)'}, alpha=0.4,
            linewidths=0.5, linecolor='gray', ax=ax);

# Dibujar puntos coloreados
for _, row in grid.iterrows():
    _ = ax.plot(row["x"] - 0.5, row["y"] - 0.5, 'o',
                color=row["region_color"], markersize=10)

# Ajustes de ejes
ax.set_xticks(np.arange(N))
ax.set_yticks(np.arange(N))
ax.set_xticklabels(np.arange(1, N + 1))
ax.set_yticklabels(np.arange(1, N + 1))
ax.invert_yaxis()
ax.set_xlabel("x")
ax.set_ylabel("y")

# Leyenda personalizada
legend_elements = [Line2D([0], [0], marker='o', color='w', label=label,
                          markerfacecolor=color, markersize=10)
                   for label, color in region_colors.items()]
_ = ax.legend(handles=legend_elements, title="Region", loc='upper left', bbox_to_anchor=(1.05, 1))

plt.tight_layout()
plt.show()
```

:::
:::
For example, part 3 of the previous example (example 1.9) can also be computed in the following way:
$$
P[X<Y] = \sum_{y=1}^{N}P[X<Y,Y=y] = \sum_{y=1}^{N}P[X<y,Y=y]
$$
Note that for $y=1$, the event $X<1$ is empty since $x \in \{1,\cdots ,N \}$. Thus:
$$
P[X<Y]=\sum_{y=2}^{N}\sum_{x=1}^{y-1}P[X=x,Y=y]=c\sum_{y=2}^{N}\sum_{x=1}^{y-1} x
$$
Using the identity $\sum_{x=1}^{y-1}=\frac{y(y-1)}{2}$, we get:
$$
=c\sum_{y=2}^{N}\frac{y(y-1)}{2}=\frac{c}{2}[\sum_{y=2}^{N} y^{2} - \sum_{y=2}^{N}y]
$$
Since:
$$
\sum_{y=2}^{N}y^{2} = \sum_{y=1}^{N}y^{2}-1 \quad \text{and} \sum_{y=2}^{N}y = \sum_{y=1}^{N}y-1
$$
we can write:
$$
P[X<Y] = \frac{c}{2}[\sum_{y=1}^{N} y^{2} - \sum_{y=1}^{N} y ]
$$
Substituting the standard formulas:
$$
\sum_{y=1}^{N-1} y = \frac{(N-1)N}{2} \quad \sum_{y=1}^{N-1} y^2 = \frac{N(N-1)(2N-1)}{6}
$$
we get:
$$
P[X<Y] = \frac{N-1}{3N}   
$$

We now proceed to extend what was done above for the case of 2 **random variables** to the case of $n$ variables.

**Definition (Discrete random vector in \mathbb{R}^{n})(por numerar)**
A random vector $(X_1,X_2, \cdots X_n)$ in $\mathbb{R}^{2}$ is said to be **discrete** if there exist a finite or countably infinite collection of vectors $(x_1^{(1)},x_2^{(1)},\cdots,x_n^{(1)}),(x_1^{(2),\cdots, x_1^{(2)}}),\cdots$ such that:
1. $P[X_1=x_{1}^{(m)},\cdots,X_n=x_{n}^{(m)}]>0 \quad \forall m$
2. $\sum_m P[X_1 = x_1^{(m)}, \cdots, X_n=x_n^{(m)}] =1$

In this case, the function
$$
f_{X_1,\cdots, X_n}\colon \mathbb{R}^{n} \rightarrow [0,1]
$$
defined by
$$
_{X_1,\cdots, X_n}(x_1,\cdots,x_n) = P[X_1=x_1, \cdots, X_n=x_n]
$$
is called the **joint density function** of the vector $(X_1,\cdots, X_n)$

The property of countable additivity implies the following relationship:
$$
F_{X_1,\cdots,X_n}(x_1,\cdots,x_n)= \sum f_{X_1,\cdots,X_n}(y_1,\cdots,y_n)
$$
More generally, if $A\subset \mathbb{R}^{n}$, the property of countable additivity property also implies:
$$
P[(X_1,\cdots, X_n)\in A] = \sum f_{X_1,\cdots,X_n}(x_1,\cdots,x_n)
$$

**Definition (Multinomial distribution)(por numerar)**
A random vector $(N_1,\cdots,N_r)$ is said to follow a **multinomial distribution** with parameters $n,p_1,\cdots,p_r$ if its joint probability mass function is given by:
$$
f_{N_1,\cdots,N_r}(n_1,\cdots,n_r) = \begin{cases}
\frac{n!}{n_1!\cdots n_r!}p_1^{n_1}\cdots p_r^{n_r} &
if \quad \sum_{k=1}^{r}n_k=n, & n_k\in \{0,1,\cdots,n\}\\
0 & otherwise
\end{cases}
$$

**Proposition (por numerar)**
Let $(N_1,\cdots, N_r)$ be a random vector with a multinomial distribution of parameters $n,p_1,\cdots,p_r$. Then, for any subcollection $N_{i_1},\cdots,N_{i_s}$ selected from thge set $\{N_1,\cdots N_r\}$, the random vector
$$
(N_{i_1}, \cdots, N_{i_s}, n - \sum_{j=1}^{s}N_{i_j})
$$
follows a **multinomial distribution** with parameters $n,p_1,\cdots,p_r, 1- \sum_{j=1}^{s}p_{i_j}$
**Proof**
First observe that the vector $(N_1,\cdots,N_r)$ follows a multinomial distribution with parameters $n,p_1,\cdots,p_r$ if and only if $N_r = n-\sum_{k=1}^{r-1}N_k$ and for any non-negative integers $n_1,\cdots, n_{r-1}$ such that $\sum_{k=1}^{r-1}n_k \leq n$, we have:
$$
P[N_1=n_1,\cdots,N_{r-1}=n_{r-1}] = \frac{n!}{n_1!\cdots n_{r-1}!(n-n_1-\cdots -n_{r-1})!}p_{1}^{n_1}\cdots p_{r-1}^{n_r-1}(1-p_1-\cdots -p_{r-1})^{n-n_1-\cdots-n_r}
$$

Nowm, it suffices to prove the result for $s=r-2$, because the general case for any $s\in \{1,\cdots, r-2\}$ can be obtained by applying the result $r-s-1$ times. Also, by reordering the collection $(N_1,\cdots,N_r)$, we may assume $i_k=k$.

Thus, for $s=r-2$ and non-negative integers $n_1,\cdots n_{r-2}$ such that $m=\sum_{k=1}^{r-2}n_k \leq n$, we have:
$$
P[N_1=n_1,\cdots,N_{r-2}=n_{r-2}] = \sum_{n_{r-1}=0}^{n-m} P[N_1=n_1,\cdots,N_{r-1}=n_{r-1}]
$$
$$
=\sum_{n_{r-1}=0}^{n-m}\frac{n!}{n_1!\cdots n_{r-1}!(n-m-n_{r-1})!}p_{1}^{n_1}\cdots p_{r-1}^{n_r-1}(1-p_1-\cdots-p_{r-1})^{n-m-n_{r-1}}
$$
Faactoring terms putside the summation:
$$
=\frac{n!}{n_1!\cdots n_{r-1}!(n-m-n_{r-1})!}p_{1}^{n_1}\cdots p_{r-2}^{n_r-2}\sum_{n_{r-1}=0}^{n-m}\binom{n-m}{n_{r-1}}p_{r-1}^{n_{r-1}}(1-p_1-\cdots-p_{r-1})^{n-m-n_{r-1}}
$$
$$
=\frac{n!}{n_1!\cdots n_{r-1}!(n-m-n_{r-1})!}p_{1}^{n_1}\cdots p_{r-2}^{n_r-2}(1-p_1-\cdots-p_{r-1})^{n-m}
$$
Which proves that $(N_1,\cdots,N_{r-2},n-\sum_{k=1}^{r-2}N_k)$ follows a multinomial distribution with parameters $n,p_1,\cdots, p_{r-2}, 1-\sum_{k=1}^{r-2}p_k$.

**Corolary (por numerar)**
Let $(N_1,\cdots,N_r)$ follows a multinomial distribution with parameters $n,p_1,\cdots,p_r$, then, for $k\in\{1,\cdots r\}$, the random variable $N_k$ follows a binomial distribution of parameters $p_k$

**Applied context**
In an electronic sensor manufacturing palnt, each sensor produced can be classified into one of the following categories after quality control:
- $e_1$:**Approved**(probability $p_1=0.8$)
- $e_2$:**Requires adjustment**(probability $p_2$=0.15)
- $e_3$:**Rejected**(probability $p_3=0.05$)

Suppose $n=20$ sensors are randomly selected. Let $(N_1,N_2,N_3)$ be the random vector counting how many sensors fall into each category.This vector follows a **multinomial distribution** with parameters $(n,p_1,p_2,p_3)$.

**Probability mass function**
The joint probability mass function of the multinomial distribution is given by:
$$
f(n_1,\cdots,n_r) = \begin{cases}
\frac{n!}{n_1!\cdots n_r!}p_1^{n_1}\cdots p_r^{n_r} & if \quad \sum_{k=1}^{r}n_k=n\\
0 & otherwise
\end{cases}
$$
**Example: Calculating the probability of specific outcome**
What is the probability that, in batch of 20 sensors, exactly 16 are approved, 3 requiere adjustment, and 1 is rejected?
$$
f(16,3,1)=\frac{20!}{16!3!1!}(0.8)^{16}(0.15)^{3}(0.05)
$$


::: {.callout-note title="Example sensor"}
#### Example sensor 
::: {.panel-tabset}
### R
```{r}
#| label: sensor-example-r
#| code-fold: true

n <- 20
p <- c(0.80, 0.15, 0.05)
outcome <- c(16, 3, 1)
prob <- dmultinom(outcome, size = n, prob = p)
round(prob, 6)
```
### Python
```{python}
#| label: sensor-example-py
#| code-fold: true
"En mantenimiento"
```

:::
:::


::: {.callout-note title="Example sensor batch 10000 sensors"}
#### Example sensor with $n=10000$ 
::: {.panel-tabset}
### R
```{r}
#| label: sensor10000-example-r
#| code-fold: true
set.seed(123)
simulaciones <- rmultinom(n = 10000, size = 20, prob = p)
df <- as.data.frame(t(simulaciones))
colnames(df) <- c("Approved", "Adjustment", "Rejected")

# Relative frequency table
tabla <- as.data.frame(table(df))
tabla$Prob <- tabla$Freq / sum(tabla$Freq)

# Visualization with ggplot2
library(ggplot2)
ggplot(tabla, aes(
  x = as.numeric(as.character(Approved)),
  y = as.numeric(as.character(Adjustment)),
  size = Prob)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    title = "Multinomial Distribution in Quality Control",
    x = "Approved Sensors",
    y = "Sensors Requiring Adjustment",
    size = "Relative Frequency") +
  theme_minimal()

```
### Python
```{python}
#| label: sensor10000-example-py
#| code-fold: true
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# 1. Fijar semilla y parámetros
np.random.seed(123)
n_simulaciones = 10000
size = 20
p = [0.80, 0.15, 0.05]

# 2. Simular datos multinomiales
simulaciones = np.random.multinomial(n=size, pvals=p, size=n_simulaciones)

# 3. Convertir a DataFrame
df = pd.DataFrame(simulaciones, columns=["Approved", "Adjustment", "Rejected"])

# 4. Calcular frecuencias relativas
tabla = df.value_counts().reset_index(name="Freq")
tabla["Prob"] = tabla["Freq"] / tabla["Freq"].sum()

# 5. Visualizar usando matplotlib
plt.figure(figsize=(10, 6))
scatter = plt.scatter(
    tabla["Approved"], tabla["Adjustment"],
    s=tabla["Prob"] * 5000,  # Escalamos tamaño para visibilidad
    alpha=0.6,
    color="steelblue"
)
plt.title("Multinomial Distribution in Quality Control")
plt.xlabel("Approved Sensors")
plt.ylabel("Sensors Requiring Adjustment")
plt.grid(True)
plt.show()
```

:::
:::

**Final reflection**
The multinoial distribution is useful in contexts such as:
- Quality of products with multiple categories
- Biostatistics (multiple clinical outcome)
- Automatic classification in engineering and machine learning
It generalizes the binomial distribution and provides a powerful tool for modeling multiple categorical outcome in repeated experiments

**Definition of density function in the n-dimensional absolutely continuous case**
The definition of a density function in the absolutely continuous case for n-dimensions is similar to the one varaible case.

**Definition (Absolutely continuous random vector)(por numerar)**
The joint distribution function
$$
F_{X_1,\cdots,X_n}
$$
of the random variables $X_1,\cdots,X_n$ is said to be **absolutely continuous** if there exists an integrable function
$$
f_{X_1,\cdots,X_n}\colon \mathbb{R}^{n} \rightarrow \mathbb{R}
$$
Such that:
$$
F_{X_1,\cdots,X_n} = \int_{-\infty}^{x_1}\cdots \int_{-\infty}^{x_n}f_{X_1,\cdots,X_n}(y_1,\cdots,y_n)dy_n\cdots dy_1 
$$
for any vector $(x_1,\cdots,x_n)$.

In this case, we also say that the random variables $X_1,\cdots X_n$ form an absolutely continuous random vector, and the function $f_{X_1,\cdots,X_n}$ is called a joint density function of $X_1,\cdots,X_n$.

Just like in the case of a single random variable, when a joint density function of $n$ random variables $X_1,\cdots,X_n$ exist, it is not unique. In fact, given one such function, we can modify its value at a finite number of points, and the new function is still a valid joint density function of $X_1,\cdots,X_n$.

Given a vector $(x_1,\cdots,x_n)\in \mathbb{R}^{n}$, define:
$$
A=\{(y_1,\cdots,y_n) \space | \space y1\leq x_1 , \cdots, y_n\leq x_n \}
$$
Then the property that charaterizes a joint density function of $X_1,\cdots,X_n$ can be expressed as:
$$
P[(X_1,\cdots,X_n)\in A] = \int \cdots \int f_{X_1,\cdots,X_n}(y_1,\cdots, y_n)dy_1\cdots dy_n
$$
it can be shown that this same relation holds for any subset $A\subset\mathbb{R}^{n}$ for which the integral
$$
\int f_{X_1,\cdots,X_{n}}(y_1,\cdots,y_n)dy_1\cdots dy_n
$$
is well defined.

**Example 1.16 (por numerar)**
Let $X$ and $Y$ be two random variables with joint density function given by:
$$
f(x,y) = \begin{cases}
\frac{1}{y}e^{\frac{-x}{y}e^{-y}} & if & x>0,y>0\\
0 & otherwise
\end{cases}
$$
Find:
(a) P[X<2Y]
(b) P[Y>2X]
**solution**


## Conditional densities

## Conditional espected value

## espected value

## Covariance and correlation

## Independence

## Espected value vector and Covariance matrix

## Normal multivariate distribution and Multinomial distribution